Ethnic and region-specific genetic risk variants of stroke and its comorbid conditions can define the variations in the burden of stroke and its phenotypic traits

Version 1

Abstract

Abstract

Burden of stroke differs by region, which could be attributed to differences in comorbid conditions and ethnicity. Genomewide variation acts as a proxy marker for ethnicity, and comorbid conditions. We present an integrated approach to understand this variation by considering prevalence and mortality rates of stroke and its comorbid risk for 204 countries from 2009 to 2019, and GWAS risk variant for all these conditions. Global and regional trend analysis of rates using linear regression, correlation and proportion analysis, signify ethnogeographic differences. Interestingly, the comorbid conditions that act as risk drivers for stroke differed by regions, with more of metabolic risk in America and Europe, in contrast to high SBP in Asian and African regions. GWAS risk loci of stroke and its comorbid conditions indicate distinct population stratification for each of these conditions, signifying for population specific risk. Unique and shared genetic risk variants for stroke, and its comorbid and followed up with ethnic specific variation can help in determining regional risk drivers for stroke. Unique ethnic specific risk variants and their distinct patterns of Linkage Disequilibrium further uncover the drivers for phenotypic variation. Therefore, identifying population and comorbidity specific risk variants might help in defining the threshold for risk, and aid in developing population specific prevention strategies for stroke.

Introduction

Stroke affects over 101 million people worldwide and is ranked the second most fatal disease in the world, with 6.5 million deaths in 20191. Comorbid conditions of stroke are critical contributors to burden of stroke and the duration of the comorbid conditions can further determine the severity of stroke risk or mortality. Prevalence for comorbid conditions range from 43% to 94% and estimates can go as high as 99% above 66 years of age2. Prevalence and mortality risk in stroke has often been evaluated from socio-economic viewpoint, but it is also critical to understand the differences in drivers such as comorbid conditions. It is the accumulated risk of comorbid conditions that enhances the risk of stroke further. Are these comorbid conditions differentially impacted by socio-economic factors and ethnogeographic factors. This was clearly evident in COVID era, when COVID-19 differentially impacted the risk of stroke, possibly due to its differential influence on the comorbidities of stroke.

Mortality in Stroke, its subtypes and their comorbid conditions have a strong ethnic bias3,4,5. Genetics act as surrogate marker for ethnogeographic indices. It is important to understand which comorbid conditions are influenced by socio-economic indices, and how they impact the risk of stroke and their underlying genetic basis. A Danish study reported the effect sizes of association with comorbid conditions for stroke to have 15% higher mortality risk in presence of diabetes mellitus with end-organ damage, 20% for peripheral vascular disease, 25% for chronic pulmonary disease, 35% for congestive heart failure and atrial flutter, 45% for moderate to severe renal disease, and 1.8- to 2.4-fold for mild to severe liver disease6. A UK Biobank study on stroke multimorbidity reported 1.5x higher risk of mortality in those with 2 additional comorbidities and a ≈2.5× higher risk of mortality in those with ≥5 comorbidities over seven years7. Thus, the differential impact of comorbid or multimorbid conditions contributing to the additive effect of illness burden needs to be addressed from an ethnogenetic perspective. Devising an appropriate strategy for prevention of stroke burden, needs a careful evaluation of the underlying genetic signature for each of these comorbid conditions and distinguishing their ethnic bias.

The objective of the study was to understand what determines the differences in stroke burden around the globe. Variations in burden of stroke could be influenced by comorbid conditions, and incidentally both stroke and its comorbid conditions can be influenced by ethnogeographic factors and genetics can act as a stable proxy marker for all. To resolve this, we considered the prevalence and mortality of a total of eleven disease conditions, comprising of stroke and its comorbid conditions, across different continents and ethnicities from 2009 to 2019. The disease conditions were further stratified as per their ethnogeographic locations and their genetic risk variants extrapolated from GWAS data. This study would provide insights on the regional patterns of the burden of stroke and its comorbid conditions, and help in resolving it from an ethnic and genetic viewpoint. These insights would further aid in developing and strategizing regional and ethnic specific needs for prevention of the risk of comorbid conditions and stroke.

Results

Global mortality, incidence and prevalence rates of stroke and its comorbid conditions

Globally, stroke ranks as the second most fatal disease in 2019 (84.2/100000, 95%UI 76.8-90.2) among the eight diseases analyzed, preceded by ischemic heart disease (IHD; 117.9/100000, 95%UI 107.8-125.9) as shown in figure S1 and table S1. High systolic blood pressure (high SBP) ranks as the most fatal comorbid disease condition with an age-standardized mortality rates (ASMR) of 138.9/100000 [95%UI 121.3-155.7] among all conditions. Within stroke subtypes, ischemic stroke (IS) ranks highest globally with ASMR of 43.5/100000 [95%UI 39.08-46.8], followed by intracerebral hemorrhage (ICH) at 36.0/100000 [95%UI 32.9-38.7] and subarachnoid hemorrhage (SAH) at 4.7/100000 [95%UI 4.1-5.2]. The ranking of the ASMRs of the diseases follow the same trend throughout the last decade with minor exceptions, where high body-mass index (high BMI) improved its rank in 2014 by swapping with high low-density lipoprotein cholesterol (high LDL), and type 1 diabetes (T1D) dethroned chronic kidney disease (CKD) in 2019.

Global trends in crude and age-standardized incidence rates (ASIRs) show that stroke incidence ranks fourth, while IHD is on top followed by T2D and CKD (fig.S1, table S1). Crude incidence rates of stroke and its subtypes increases in the last decade but ASIRs decreases, with exception of IS where an increase in ASIR was observed. Among other diseases, IHD, type 2 diabetes (T2D) and T1D show a continuous increase in the last decade, with T2D (280.1/100000, 95%UI 258.8-303.9) surpassing IHD (274.0/100000, 95%UI 242.9-306.4) in 2019 in crude rates.

The global trend between crude and age-standardized prevalence rates (ASPRs) revealed that the ranking of stroke remains at sixth position throughout the years (fig.S1, table S1). Among stroke subtypes, IS ranks higher over the ICH and SAH. The comorbid conditions, high SBP and high BMI, ranks first and second followed by CKD and T2D, and interestingly all rank above stroke. We also observe that though there is a continuous increase in prevalence rates in the last decade, the ranking of stroke or its comorbid conditions does not change over the years, with the exception of T1D ASPRs overtaking ICH. We were keen to understand if these global trends of ASMR and ASPR are influenced by region and ethnicity.

Ethno-regional differences in mortality and prevalence of stroke and its major comorbid conditions

We observed interesting patterns of ASMRs of stroke, its subtypes and its major comorbidities across different regions over the years as shown in figure 1a, S3, table 1 and S2. When assessed in terms of ranks, high SBP is the most fatal condition followed by IHD in all regions, except Oceania (OCE) where IHD and high SBP swap ranks. Africa (AFR; 206.2/100000, 95%UI 177.4-234.2) and Middle East (MDE; 198.6/100000, 95%UI 162.8-234.4) have the highest ASMR for high SBP, even though they rank as only the third and sixth most populous continents (fig. S2), respectively. Both high SBP (-0.64% to -2.25% EAPC) and IHD (-0.45% to -1.17% EAPC) show a decreasing trend for ASMRs in all regions. However, only Europe shows a significant decrease for high SBP (-2.26%; P=0.009) and IHD (-2.37%; P=0.006) in the decade. Stroke has a decreasing trend for ASMRs with East Asia (EAS) and Europe showing a significant decrease of -2.2% (P=0.021) and -2.6% (P=0.03), respectively. Stroke has the highest mortality in EAS (127.1/100000, 95%UI 104.9-150.5 in 2019), and is the only region that ranks stroke higher than IHD. Though EUR, MDE and Central & South Asia (CSA) have ASMRs similar to global rates for stroke, CSA ranks stroke as the third most fatal factor, while America (AMR), Europe and MDE ranks it fifth. Oceania (62.1/100000, 95%UI 34.1-90.2) and America (40.3/100000, 95%UI 36.2-43.1) have lowest rates for stroke in 2019.

Regional (a) mortality rates and (b) prevalence rates for Stroke, its subtypes and comorbid factors.

The figure shows the age-standardized mortality and prevalence rates per 100,000 people for Stroke, its subtypes and its comorbid factors in 2009, 2014 and 2019. The size of the points indicates the rate and position indicates rank.

Among the stroke subtypes, ICH and IS show maximum ethnogenetic differences in mortality rates (14.1/100000 to 61.4/100000) and ranking (4th to 9th) in 2019. While ICH shows a significant decrease in ASMR in EAS (-3.53%, P=0.009), IS shows a significant decrease in Europe (-2.37%, P=0.06). High BMI and high LDL rank in top five but their mortality rates differed across all regions, with the highest rates for both in MDE. Only Europe shows a significant decrease in high LDL (-2.53%, P=0.03) over the decade. MDE has the highest ASMRs due to IHD and high SBP, followed by AFR, CSA, EAS and EUR, all having rates higher than global. All continents have similar mortality rates for T2D and CKD across the years, except Oceania, where the T2D rate is nearly three times CKD rate. Africa has the highest mortality rate for T1D (1.59/100000, 95%UI 1.2-1.9).

ASPRs also showed an interesting pattern of distribution, and, in contrast to mortality, show an increase over the decade (fig. 1b and S4, table 2 and S3). Highest ASPRs were observed for high SBP across all regions, except America, MDE and Oceania, where high BMI has most prevalence. While EAPC of high SBP showed significant decrease in all regions (-0.38% to -1.77%), except CSA, high BMI showed a significant increase (1.77% to 5.6%) in all (table 2). The ASPR ranking of CKD and T2D rose to top five, in sharp contrast to their ASMR rankings. Prevalence of CKD (EAPC 0.24% to 0.7%) and T2D (EAPC 0.6% to 2.18%) is significantly increasing in all regions. For all other diseases, the pattern of ranking and rates across regions were stable with minor exceptions. Stroke ranks sixth for ASPRs in all regions, and it is interesting to note that ASPRs of all the comorbid conditions, except T1D, rank above stroke. While Europe shows a significant decline in ASPRs of stroke (-0.9%, P=0.008) and IS (-0.85%, P=0.03) across the years, EAS shows a significant increase for stroke (0.7%, P=0.02) and IS (1.09%, P=<0.001). Globally, T1D swapped its ASPR ranking with ICH in 2014, largely influenced by the significant increase in ASPRs in MDE (2.83%, P=<0.001). However, the highest prevalence of T1D is in Europe and Oceania. The prevalence of IHD has remained nearly constant in all continents in the last decade, except Oceania (-0.43%), America (-0.49%) and MDE (-0.27%) which shows a significant decrease. In 2019, MDE has the highest prevalence for IHD (4843.02/100000, 95%UI 4243.02-5442.58), while America (1695.6/100000, 95%UI 1530.8-1871.9) has the lowest ASPR, less than half the rate of the MDE. Globally, T1D swapped its ASPR ranking with ICH in 2014, largely influenced by the significant increase in ASPRs of T1D in MDE (2.8%, P=<0.001). However, the highest prevalence of T1D is in Europe and Oceania. We were further keen to understand if these regional differences in ASMR and ASPR also reflect a socio-economic bias and if so, does it reflect in a category of comorbid conditions.

Contribution of metabolic risk and hypertension in stroke based on ethnogenetic locations

When the prevalence and mortality rates of stroke and its comorbidities were grouped into three groups, namely strokes, metabolic disorders and high SBP, we find that out of the three proportional mortalities shown in figure 2, strokes group has the highest proportion (37.1%-47.2%) across all years and regions, except Oceania and America, where instead, metabolic disorders have the highest proportion (39.1%-42.2%) that is significantly higher compared to global proportion (table S4). East Asia has the highest proportional mortality for strokes among all regions in all three years (44.7%-47.2%). This was in sharp contrast to prevalence proportion of strokes (4.6%-9.1%), which was the least among the three groups, with the highest proportional prevalence for strokes being 8.9%-9.1% in CSA. The proportional mortality for high SBP (22.0%-30.5%) is very similar across the regions. Metabolic disorders have significantly higher proportional prevalence compared to global in MDE, Oceania and America, the highest being in America (63.9%, table S5). Asian and African regions have lowest proportional prevalence for metabolic disorders, with CSA having significantly lower proportion. However, these regions have the highest prevalence proportion for high SBP (46.6% - 54.3%), with CSA having a significantly highest proportion compared to global. On the other hand, MDE, Oceania and America has significantly lower proportional prevalence compared to global. We were further keen to understand the correlation among ASMR and ASPR of comorbid conditions among ethnogeographic regions.

Proportional (a) mortality rates and (b) prevalence rates for Strokes, High SBP and Metabolic disorders.

The figure shows the proportional mortality and prevalence for all strokes in yellow (ischemic stroke, intracerebral hemorrhage, subarachnoid hemorrhage and ischemic heart disease), high systolic blood pressure in red, and metabolic disorders in green (high BMI, high LDL, diabetes mellitus type 1 and 2, chronic kidney disorder). CSA - Central and South Asia, AFR - Africa, EAS - East Asia, EUR - Europe, MDE - Middle East, OCE - Oceania, AMR – America.

Correlation among prevalence and mortality rates based on ethnogeographic region

Correlation between ASMRs and ASPRs for stroke and all comorbid conditions across each ethnogeographic locations is shown in figure 3. High SBP prevalence and mortality show a strong positive correlation in CSA but a strong negative correlation in MDE. The prevalence and mortality of high BMI has a strong negative correlation in CSA and MDE populations, but a strong positive correlation in EAS. Prevalence of CKD has negative correlation with mortality rates in EAS. Prevalence of T2D has negative correlation with mortality rates in Oceania and positive correlation in America. It is interesting to note that there was not much of a correlation in mortality and prevalence rates for most of the conditions. For overall stroke, though minor correlations between various ethnicities are seen, this becomes alarmingly clear in the stroke subtypes. The correlation matrix of ASMR and ASPR of stroke and its comorbidities does reflect strong ethnogeographic distinctions, which formed the basis of further investigation on genetic basis of stroke and its comorbidities.

Correlation of mortality versus prevalence rates of Stroke, its subtypes and comorbid factors.

The figure shows the Pearson correlation coefficients of mortality rates versus prevalence rates across different continents for 2014. CSA - Central and South Asia, AFR - Africa, EAS - East Asia, EUR - Europe, MDE - Middle East, OCE - Oceania, AMR - America.

Ethnogeographic stratification of stroke and its comorbidities based on GWAS data

To resolve the ethnogeographic distinctions for stroke and its major comorbid conditions based on their genetic risk, we considered all GWAS loci for stroke and its major comorbid conditions, and subjected it to stratification analysis. From the GWAS loci, we observed a distinct population structure that distinguished ethnogeographic populations based on their genetic signatures (Fig. 4). For all diseases, except high BMI, the individuals clustered into five groups, each corresponding with the five super-populations from 1000 Genome project namely, African, East Asian, South Asian, European and American. For high BMI, the individuals clustered into three groups corresponding to African, East Asian, European. Though broad clustering of ancestral populations among the diseases looks similar, the proportions of ancestral populations in certain diseases vary greatly. Among Stroke, IHD and T2D risk variants, the populations structured in a similar way, while for T1D, CKD and LDL the patterns were slightly different in European and South Asian ancestry. Whereas for high SBP the major fluctuations were observed in South Asian and East Asian populations. This stratification was further explored using population-based clustering, where a similar pattern was observed in PCA plots for stroke and its comorbid conditions (Fig. 5). African population seems to be a distinct outlier for most diseases, and the East Asian comes a distant second in the cluster pattern.

Population structure of risk variants for Stroke and its comorbid factors using a model-based clustering.

The proportion of ancestral populations was estimated from the genotype of risk variants of each disease in unrelated individuals from the 1000 Genomes project using a model-based clustering approach. The individuals were represented by their super-population in 1000 Genomes (African, East Asian, South Asian, European and American). The estimated ancestral populations clustered into either 5 clusters or 3 clusters (in case of BMI) represented by the different colors.

Clustering of risk variants for Stroke and its comorbid factors using a PCA.

The eigenvectors were estimated from the genotype of risk variants of each disease in unrelated individuals from the 1000 Genomes project using PCA clustering. The individuals were represented by their super-population in 1000 Genomes (African, East Asian, South Asian, European and American) shown in five different colors.

Similar patterns were observed in stratification analysis after grouping the GWAS loci of stroke and its comorbidities into three groups as before, namely strokes, metabolic disorders and high SBP. The African, East Asian and South Asian populations had distinct structure in all three groups (fig. S5). These observations do indicate that the underlying genetic factors of stroke and its comorbid factors can be the real indicators of ethnogeographic patterns of risk for stroke and its comorbid conditions. However, we were further keen to understand the extent of shared and unique individual risk variants across stroke and its comorbid condition and how these unique or shared variants can help in distinguishing their relevance across ethnicities.

Shared and unique risk variants of stroke among the different ethnogenetic regions

The unique and shared individual variants across stroke and its comorbid conditions were identified from the GWAS data irrespective of ethnicity. We find majority of the risk variants were unique to a disease condition, however, several risk variants were also seen to be shared with stroke and other comorbid conditions as seen in Figure 6. We were further keen to have a deeper insight into the distribution of risk variants in stroke across ethnicities. Stroke has only 55% of the risk variants common to all the five populations as seen in figure 7 and table S6. Two groups of populations share the most number of variants, namely, the Africa-America-Europe-South Asia (6% of variants shared) group and the East Asia-America-Europe-South Asia (4%) group. Africa has the highest number of unique variants for stroke (6%), followed by Europe (3%).

Distribution of unique and shared risk variants in stroke and its comorbid conditions.

The barplot shows the number of GWAS risk variants unique for stroke and its comorbities, as well as number of GWAS risk variants shared among the different diseases. The intersection of diseases is indicated below. Intersections containing stroke are highlighted in red. The total number of GWAS risk variants considered are stroke (366), IHD (1137), High SBP (419), T2D (2227), T1D (269), CKD (125), High BMI (27) and High LDL (1734).

Risk variants of Stroke shared among super-populations.

The GWAS risk variants for stroke (total 291) shared among the five super-populations in 1000 Genomes (African, East Asian, South Asian, European and American). A risk variant was considered to be present in a population if the alternate allele frequency in 1000 Genomes was greater than or equal to 0.05.

South Asia has two unique variants for stroke, rs528002287 and rs148010464, which maps to genes PCSK6 and the intergenic region of PLA2G4A/LINC01036, respectively. The variants rs528002287 and rs148010464 are low frequency variants in South Asia with a MAF of 0.053 and 0.054, respectively. We were further keen to understand how these unique variants in South Asia are tagged to the nearby variants of different frequencies in the different ethnicities. Can the LD patterns of rare allele and common allele help in distinguishing the ethnogeographic distinction in phenotype variation. On comparison, the linkage disequilibrium (LD) pattern of low frequency variants and common variants across ethnicities showed contrasting patterns. The LD plots between the unique variants and low frequency variants tagged to it clearly demonstrated a unique LD pattern in South Asia, compared to other populations (Figure 8). Contrastingly, the LD plots of common variants tagged to the risk variant of stroke unique to South Asia showed similar LD patterns among all populations (fig. S6). These differences might also reflect unique or distinct phenotypic differences among ethnicities for risk in stroke.

Linkage disequilibrium of low frequency variants proxy to risk variants of stroke unique in South Asian population.

Proxy variants with frequency less than 0.1 in 1MB region flanking the stroke risk variants unique in South Asia (a) rs528002287 and (b) rs148010464. The risk variants are marked in blue box. LD plots of variants in different 1000 Genome super-populations (AFR- Africa, EAS- East Asia, SAS – South Asia, AMR – America, EUR – Europe) are shown.

Discussion

To the best of our knowledge, this is the first study that explored the burden of stroke and its comorbid conditions across regions, stratifying and distinguishing their unique features based on their genetic background extrapolated from the GWAS risk loci. The dynamics of different rates of stroke, its subtypes and comorbid factors do reflect ethnogeographic differences. Globally, the prevalence and incidence rates of stroke have increased, while mortality rates decreased with minor shifts in ranking in the last decade. Interestingly, the incidence and prevalence rank of stroke rates remain the same globally, but for mortality it ranks third, preceded only by IHD and high SBP. While the global stroke prevalence is nearly 15 times its mortality rate, prevalence of comorbid conditions such as high SBP, high BMI, CKD, T2D are alarmingly 150- to 500-fold higher than their mortality rates. These comorbid conditions can drastically affect the outcome of stroke. Interestingly, these disparities in rates get further widened when evaluated from an ethnogeographic perspective. The age-standardized prevalence of stroke in 2019 ranges from lowest in CSA (858.5/100000, 95%UI 737.6-979.4) to highest in EAS (1513.1/100000, 95%UI 1390.7-1635.5), in contrast to mortality rates, lowest in America (40.3/100000, 95%UI 36.2-43.1) and highest in EAS (127.7/100000, 95%UI 104.9-150.5).

The rates of stroke, its subtypes and comorbid conditions do correlate to some extent but their ranking varies significantly. In terms of ranking, stroke ranks sixth in prevalence across all ethnogeographic locations, but ranks second in mortality in EAS and third in CSA and Africa. We find that among the considered comorbid conditions, some rank above stroke in both incidence and prevalence, however in terms of mortality, stroke ranks highest with exception to high SBP. Similarly, the ranking of the comorbid conditions also varies when the global population is stratified based on ethnogeographic locations. In the last decade, there has been tremendous development in the healthcare industry globally, but this is not reflected in the mortality or prevalence data from 2009-2019. Ranking of comorbid conditions by the rates is very crucial to identify ethnic-specific comorbid risk that can be helpful in guiding and managing stroke risk.

The changing dynamics of stroke or its comorbid conditions can be attributed to multitude of factors. Often global burden of stroke has been discussed from the point of view of socio-economic parameters. Studies indicate that half of the stroke-related deaths are attributable to poor management of modifiable risk factors8,9. However, we observe that different socio-economic regions are driven by different risk factors. Considering that Europe, America, Oceania and the Middle East represent high socio-economic regions, the comorbid conditions that drive prevalence and mortality rates here seem to be more of metabolic in nature, while for South Asians, high SBP is the prominent factor. It is evident from the correlation between prevalence and mortality for stroke, its subtypes and comorbid conditions that there is an ethnic and comorbid specific correlation, which possibly does not reflect a clear socio-economic distinction. The comorbid conditions for stroke subtypes also differ to a large extent. Diabetes is a comorbid condition for stroke but not for SAH, and this risk is ethnic-specific10,11. Therefore, it is very pertinent to understand the stroke risk from an ethnic view point, beyond the boundaries of socio-economic criteria, as the drivers of comorbid risk and ethnicity rely on genetic and epigenetic components. Studies reported reduction in life expectancy in 31 of 37 high-income countries, deduced to be due to COVID-1912. However, it would be unfair to ignore the comorbid conditions which could also be the critical determinants for reduced life expectancy in these countries.

Stroke has a complex etiology, which is further influenced by its comorbid conditions and this impacts its phenotypic variability. A strong genetic risk drives both stroke and its comorbid conditions. Genetic risk variants for diabetes, cardiovascular disease, diabetic retino- and nephropathies, hypertension, inflammation and kidney diseases have been reportedly shown to have strong ethnogenetic variation13. Implications of ethnogenetic differences were evident when GWAS genes for stroke and all studies comorbid conditions were used to stratify the 1000 genome super-populations. We observed that the GWAS risk loci for stroke and its comorbid conditions like high BMI, high LDL, high SBP, T2D, T1D, and CKD could stratify the super-populations based on its ethnogenetic considerations. Fluctuations in genetic structure of stroke and its comorbid conditions signify the impact of ethnic variations on mortality and prevalence rates. Stroke accounts for approximately 20% of deaths in diabetics14,15. Diabetics and pre-diabetics, and the duration of diabetes have been reported to have increased risk of stroke, which gets aggrevated in African-Americans14,15. As stroke and its comorbid conditions are heavily influenced by lifestyle, high-income countries showed evidence of metabolic disorders being the major cause of concern for both prevalence and mortality. Similar observation in a UK biobank cohort study on stroke suggests that genetic and lifestyle factors were independently associated with incident stroke, which emphasizes the benefit of entire populations adhering to a healthy lifestyle, independent of genetic risk16.

Metabolic risk variations could also be a reflection of their underlying genetic differences. Significant differences among ethnicities in metabolism of various macromolecules have been reported17. Genetic variations demonstrate inter-ethnic differences in LDL levels resulting in differential impact on dyslipidemia18. A meta-analysis on European, East Asian and African-American ethnicities revealed that common variants of CDH13 and ADIPOQ regulate adiponectin levels, an important component of BMI indicator19. ALDH2*504Lys allele has been reported to be associated with high BMI, increased tolerance of alcohol, high SBP, and decreased high-density lipoprotein (HDL) in East Asians20,21. MEGASTROKE consortium on stroke and its comorbid factors reported five variants associated with blood pressure, two with LDL cholesterol and reported that all stroke subtypes were associated with a Genetic Risk Score (GRS) for high SBP22. A study on IHD, T2D in European populations with different health-care systems, and local population substructures reported Polygenic risk score (PRS) with similar accuracy across Europeans, to a lesser extent to South and East Asian populations, and very poor transferability for Africans23.

Even stroke subtypes show a strong ethnic variation22,24. While COL4A1 and COL4A2 were common denominators for most of the stroke subtypes, HDL was inversely associated with small vessel stroke25. The INTERSTROKE study on stroke subtypes reported that among comorbid factors, high SBP was significantly associated with ICH and diabetes, cardiac issues, and apolipoproteins with ischemic stroke26. Ischemic stroke associated with ApoB/ApoA1 ratio had higher (67.6%) population attributable fraction (PAF) in Southeast Asia compared to Western Europe, North America, and Australia (24.8%), while ischemic stroke associated with atrial fibrillation had lower PAF in South Asia (3.1%) compared to the rest (17.1%)26. Emerging studies like novel ethnic specific genetic variants, such as SUMOylation pathway in Indians27, SFXN4 and TMEM108 in Africans,28 indicate the involvement of different pathways among different ethnicities in stroke. The A allele of c.*84G>A loci in CETP gene was found to be a risk factor for IHD in South Asians29. High SBP was found to be a risk factor in all major stroke subtypes except lobar ICH30. Therefore, identifying differential risk in different ethnicities for stroke and its subtypes, and its impact on comorbid conditions might also indicate different treatment modalities which can minimize adverse metabolic side effects.

Identifying the pattern of genetic variation is critical in distinguishing stroke and its endophenotypic variations. The risk variants rs528002287 (locus 15q26.3) in PCSK6 and rs148010464 (locus 1q31.1) an intergenic variant in PLA2G4A/LINC01036 for stroke were unique to South Asia, and were found to be associated with cardioembolic stroke and small vessel stroke in South Asians27. A recent INTERSTROKE study reported association of short sleep duration with increased risk for stroke to be highest in the South Asian ethnicity (OR 9.13, 95% CI 5.86-14.66)43. Decreased sleep quantity and quality has been reported to increase blood pressure44, prevalence for which was found to be highest for South Asians in our study. These observations are interesting as PCSK6 is known to regulate sodium homeostasis and thereby maintaining diastolic blood pressure.31,32 Reports also indicate 2.34 fold difference in the expression of PCSK6 during maintenance phase of hypertension33. PCSK6 has also been reported to be involved in processing of precursors of Melanin Concentrating hormone (MCH) under certain conditions45. MCH is known to play a central role in promoting and stabilizing sleep46,47. In insomnia patients, PLA2G4A was reported to be upregulated by 1.88 fold after improvement in sleep48. In sleep deprived mice, glycolytic pathway and lipid metabolism was upregulated and expression of PLA2G4A was downregulated49. These observations are interesting as PLA2G4A is known to play a role in the metabolism of phospholipids, production of lipid mediators and the release of arachidonic acid (AA)35,36. AA is involved in signaling pathways of metabolic processes like release of insulin and glucose disposal37-39. PLA2G4A also plays a role in the production of pro-thrombotic TXA2 and thus, inhibition of PLA2G4A can reduce platelet aggregation and thromboembolism40. Thus, the risk of unique genetic variants in PCSK6 and PLA2G4A in South Asian ancestry may indicate a unique endophenotype for stroke, which might also indicate the influence of underlying risk variants for comorbid conditions. E.g. PLA2G4A in metabolic processes.

GWAS has yielded numerous common risk alleles that are associated with various human phenotypes50,51. Since the rationale for GWAS is the ‘common disease, common variant’ hypothesis, it has been able to identify only those common variants with a moderate effect on the associated trait and thus, these identified variants only explain a small proportion of the heritability of the trait. One of the major conclusions from the 1000 Genomes Project was that most variations in the human genome are rare and unique to specific sub-populations52,53. From an evolutionary point of view, alleles with strong effects that are detrimental will be controlled by purifying selection keeping its frequency low. Hence, rare and low frequency variants might be variants with large effects on traits54. Examples of genes like ABCA1, PCSK9 and LDLR55,56, which carries both common variants with moderate effects as well as rare variants with large effect for lipid levels indicate that genes can contain both types of variants associated with a complex trait. Using this logic, we were keen to identify the influence of common and rare variants in stroke and its comorbid conditions and their pattern of LD in distinguishing ethnic specific risk.

While we observe majority of the GWAS variants associated with stroke are common variants, a minority of these are low frequency variants with below 10% frequency. The alternate alleles of these variants were found to be present only in specific super-populations of 1000 Genomes, while the variants are monoallelic in the other super-populations. This difference among populations gets further enhanced when we look at the LD patterns of these low frequency variants with other low frequency variants which are in proxy. Distinct LD blocks with these unique rare variants present in one super-population was seen to be absent in other populations. On the other hand, LD patterns of common variants (frequency greater than 10%) in the same regions shows similar LD patterns for all populations. Thus, the rare variant hypothesis could explain a significant proportion of the differences in burden of stroke seen across populations. Such genes identified could be possible candidate genes for identifying rare and low frequency variants that could play a role in the heritability of stroke and its comorbid factors.

Conclusions

The dynamics of incidence, prevalence and mortality rates of stroke and its subtypes along with its comorbid risk factors reflect strong ethnogeographic differences. Our work highlights that these ethnogeographic differences for stroke and its comorbid conditions need to be evaluated and stratified based on their their ethnogenetic background. Genetic variables should be considered as primary evidence as they define the threshold for biochemical, metabolomic or epigenetic variables. The different socio-economic regions are driven by different risk factors of stroke and low frequency variants could be playing a role in the differences in burden of stroke seen in the different regions. Identifying population specific unique variants for stroke and its comorbid conditions might refine the drivers for endophenotypic variations for stroke risk. We would like to suggest that integrating public health genomics and articulating it with comorbid conditions for stroke should be considered crucial irrespective of the economic status, as both lower and higher socio-economic regions have different drivers of stroke risk.

Methods

Data sources

We obtained age-standardized incidence rates (ASIRs), age-standardized prevalence rates (ASPRs) and age-standardized mortality rates (ASMRs) for a total of eight diseases and three disease conditions in 204 countries, for the years 2009 to 2019 using the GBD Results Tool (https://vizhub.healthdata.org/gbd-results/)58 and from NCD Risk Factor Collaboration (NCD-RisC, https://ncdrisc.org/) study59,60 in November 2021. GBD codes for the selected diseases were B.2.2 (Ischemic heart disease, IHD), B.2.3 (Stroke), B.2.3.1 (Ischemic stroke, IS), B.2.3.2 (Intracerebral hemorrhage, ICH), B.2.3.3 (Subarachnoid hemorrhage, SAH), B.8.1.1 (Diabetes mellitus type 1, T1D), B.8.1.2 (Diabetes mellitus type 2, T2D) and B.8.2 (Chronic kidney disease, CKD). Three disease conditions, which are also comorbid factors of stroke, were also selected, high systolic blood pressure (SBP) (>110-115 mmHg), high body mass index (BMI > 23.0 kg/m2), and high LDL cholesterol. For these, ASMRs in 204 countries, as well as global rates for 2009 to 2019 were obtained using the GBD Results Tool, and age-standardized prevalence percentages in 204 countries for the years 2009 to 2016 (data for 2017 to 2019 was not available) were obtained using NCD-Risc website (https://ncdrisc.org/)59,60. Global crude rates and age-standardized incidence, prevalence and mortality rates were obtained from GBD and NCD for all the diseases and disease conditions (Figure S1, Table S1). GBD 2019 and NCD-RisC study compiled with the GATHER Guidelines.

Spatio-temporal trend analysis and estimated annual percent change

The 204 countries were grouped into eight geographic regions namely Global, America (AMR), Europe (EUR), Middle East (MDE), Africa (AFR), Central Asia & South Asia (CSA), East Asia (EAS), and Oceania (OCE), based on ethnicity (Table S7, Figure S2). The prevalence percentages were converted into ASPRs for each disease condition. ASMRs and ASPRs for each region was obtained using Bayesian model averaging of linear regression models with Markov chain Monte Carlo sampling. BIC was the model selection criteria. Poisson distribution with global rates as lambda was the prior distribution on the models. The population of each country was used as weights. 10000 draws from the posterior distribution of model parameters were used to obtain the point estimates (mean of the draws) and 95% uncertainty intervals (2.5th to 97.5th percentiles of the posterior distributions) of mortality and prevalence rates for each region for years 2009 to 2019. Rate estimates presented in this paper are age-standardized rates per 100,000 population.

The ASMRs and ASPRs thus obtained for the different diseases in each region were subjected to a temporal rank analysis for 2009, 2014, and 2019 using custom perl scripts and change in the trend was plotted as a bump plot. The size and position of the points indicates the rate and rank respectively. To quantify the temporal trends, estimated annual percentage change (EAPC) was modeled using Poisson regression using a generalized linear model for the log-transformed rates: log(y)= β0+ β1x1 + β2x2 + ….+βpxp, where y is the age-standardized rate, xi are the calendar years, βi are the rate trends. Under the assumption of linearity of log of age-standardized rate with time, EAPC =100* exp(β^) – 1. The 95% uncertainty interval is calculated as CI(EAPC) = β + (Z(1-α)/2) x SE, where α is the confidence level, SE is the standard error of β^.

EAPC for ASMRs and ASPRs were calculated for the time period 2009 to 2019 (table 1 and 2). EAPC for high SBP prevalence was calculated from 2009 to 2015 and for high BMI prevalence from 2009 to 2016. EAPC was considered statistically significant if the uncertainty interval of EAPC did not cross zero. Statistical significance of spatio-temporal difference in rates was calculated using chi-square test by assuming the rate to be under Poisson distribution. 2019 ASMRs (or ASPRs) in each region were compared with global ASMR (or ASPR) to compare change over locations and with 2009 ASMRs (ASPRs) to compare change over time. 2014 ASPRs were used when 2019 data was not available. P-values (two-sided) for ASMRs and ASPRs are shown in table S3 and S4, respectively. Chi-square tests were done using Open Source Epidemiologic Statistics for Public Health (www.openepi.com/Menu/OE_Menu.htm). The relation between ASMRs and ASPRs of each region was measured using Pearson correlation. For correlation, we considered the data of 2014 due to the complete spectrum of data availability. All analysis, unless specified, was done using R Statistical Software (version 4.1.2) with packages BAS, Rcan, corrplot, dplyr and ggplot2.

Proportional mortality and prevalence

The diseases were classified into three categories as stroke (IS, ICH, SAH and IHD), metabolic disorders (High BMI, High LDL, T2D, CKD, and T1D) and high SBP. To determine the proportion of each category in a region, total ASMRs were scaled to 100 for all three years. The same was done for ASPRs (high LDL cholesterol data was not available). Statistical significance of difference in proportions was calculated using one-sample test for binomial proportion using normal-theory method. The proportion mortality in each category was compared in a pairwise manner with global as well other regional proportions to calculate the two-sided p-value. The same was done for proportional prevalence. P-values for proportional mortality and prevalence are shown in tables S5 and S6, respectively. Proportion comparison were done using Open Source Epidemiologic Statistics for Public Health (http://www.openepi.com/Menu/OE_Menu.htm).

Population Structure Analysis

For evaluating the ethnogenetic perspective of stroke and its comorbid conditions we considered the risk variants associated with each disease. The risk variants were obtained from GWAS Catalog (https://www.ebi.ac.uk/gwas/home), during the period November 2021 to August 2022. The trait IDs used to retrieve data from GWAS Catalog for the different diseases were EFO_0000712 (Stroke), EFO_0001645 (IHD), EFO_0000537 (High SBP), MONDO_0005148 (T2D), MONDO_0005147 (T1D), EFO_0003884 (CKD), EFO_0007041 (High BMI) and EFO_0004611 (High LDL). The position of the risk variants in GRCh37 assembly was obtained, and variants less than 10000 bp apart were excluded using custom perl scripts. The total number of risk variants for each disease thus obtained are shown in table S8. The genotype of identified biallelic autosomal SNPs in unrelated individuals was extracted from the 1000 Genome VCF files (https://www.internationalgenome.org/). The proportion of ancestral populations in each individual was estimated from their genotype using a model-based clustering approach61. The admixture model with correlated allele frequencies was specified to cluster the individuals into either five clusters or three clusters (in case of BMI). The genotype was converted to eigenvectors using principal component analysis62,63.

Shared and Unique risk variants among ethnicities

The number of GWAS risk variants for stroke shared among, as well as, unique to the five super-populations in 1000 Genomes (African, East Asian, South Asian, European and American) was obtained (Fig.6 and Table S6). A risk variant was considered to be present in a population if the alternate allele frequency in 1000 Genomes was greater than or equal to 0.05. For each GWAS risk loci, the gene(s) the variant maps to was obtained and shared genes among the populations was also estimated (table S6).

Calculation of Linkage Disequilibrium (LD) of risk variants of stroke in South Asia

Proxy variants (R2 > 0.01) in the region -/+ 500 Kb of the risk variant of stroke present in South Asian population was obtained from LDlink64. Among the proxy variants, variants with minor allele frequency (MAF) less than 0.1 were selected along with the risk variant as low frequency variants, and variants with MAF greater than 0.1 were termed as common variants. LD between two alleles A and B is quantified using the coefficient of linkage disequilibrium DAB calculated using the equation DAB = pAB-pApB, where pi represents the frequency of the allele i or haplotype i. To be able to compare the level of LD between different pairs of alleles, D is normalized as follows: D’ = D/ Dmax, where Dmax = max{-pApB, -(1-pA)(1-pB)} when D < 0 and Dmax = min{pB(1-pA), pA(1-pB)} when D > 065. Estimates of D’ were calculated separately for low frequency variants and common variants, and plotted using the R library gaston.

eLife Assessment

This paper provides a useful analysis of the variation of the burden of strokes across geographic regions, finding differences in the relationship between strokes and their comorbidities. This dataset and the correlations found within will be a resource for directing the focus of future investigations. The statistical analyses are incomplete.

Reviewer #1 (Public Review):

Summary:

The paper measures the prevalence and mortality of stroke and its comorbidities across geographic regions in order to find differences in risks that may lead to more effective guidance for these subpopulations. It also does a genetic analysis to look for variants that may drive these phenotypic variations.

Strengths:

The data provided here will provide a foundation for a lot of future research into the causes of the observed correlations as well as whether the observed differences in comorbidities across regions have clinically relevant effects on risk management.

Weaknesses:

As with any cross-national analysis of rates, the data is vulnerable to differences in classification and reporting across jurisdictions. Furthermore, given the increased death rate from COVID-19 associated with many of these comorbid conditions and the long-term effects of COVID-19 infection on vascular health, it is expected that many of the correlations observed in this dataset will shift along with the shifting health of the underlying populations.

Reviewer #2 (Public Review):

Summary:

The authors have analyzed ethnogeographic differences in the comorbidity factors, such as diabetes and heart disease, for the incidences of stroke and whether it leads to mortality.

Strengths:

The idea is interesting and the data are compelling. The results are technically solid.

The authors identify specific genetic loci that increase the risk of a stroke and how they differ by region.

Weaknesses:

The presentation is not focused. It would be better to include p-values and focus presentation on the main effects of the dataset analysis.

Author response:

The following is the authors’ response to the current reviews.

eLife assessment: I find that the eLife assessment mentions “statistical analyses are yet to be carried out to support statements of statistical significance” while the reviewers mention that the data are compelling and results are technically solid. Besides all observations in the manuscript are presented with robust and established norms of statistical analysis.

Public Reviews:

Reviewer #1 (Public Review):

Strengths:

The use of data from before COVID-19 is both a strength and a weakness. Because COVID had effects on vascular health and had higher death rates for groups with the comorbidities of interest here, it has likely shifted the demographics in ways that would shift the results in unpredictable ways if the analysis were repeated with current data. This can be a strength in providing a reference point for studying those changes as well as allowing researchers to study differences between regions without the complication of different public health responses adding extra variation to the data. On the other hand, it limits the usefulness of the data in research concerned with the current status of the various populations.

We completely agree with the observation, but were restricted as the purpose was to use the most robust and technically qualified data from GBD. The post COVID19 GBD data has not yet been released, but I am sure the observations made in the study can help in guiding the issues in the post COVID era too, because genetics is not going to change in these population groups.

However, we did highlight this aspect of COVID19 even in our original version and also in the revised version.

Reviewer #2 (Public Review):

Weaknesses:

The presentation is not focused. It is important to include p-values for all comparisons and focus the presentation on the main effects from the dataset analysis.

The significant p-values were restricted to public health data only to identify and distinguish differences in incidence, prevalence and mortality and how they differ across world populations. These differences have often been interpreted from socio-economic point of view, while our manuscript presents the reasons for differences for main condition (Stroke) and its comorbid condition among different ethnicities from a genetic perspective. This genetic perspective was further explored to identify unique ethnic specific variants and their patterns of linkage disequilibrium in distinguishing the phenotypic variations. Considering the quantum and diversity of data, both for public health and GWAS data, there can be several directions but for presentation we focused only on the most distinguishing and established phenotypic differences. I am sure this will open up avenues for several future investigations including COVID, as has been highlighted by the reviewers too. All observations were presented with robust and established norms of statistical analysis.

The following is the authors’ response to the original reviews.

Thanks for the constructive observations on strengths and weaknesses of our manuscript. Interestingly, some of the weaknesses mentioned here also turns out to be the strength of the article. For example COVID19 has been mentioned by the reviewer as a driver to increase the mortality in some comorbid conditions and stroke. Firstly, I must clarify that, our data is from PreCOVID era and we indeed mention that in COVID era, COVID-19 might differentially impact the risk of stroke. Possibly this differential influence on the comorbidities of stroke, is likely to be influenced by its underlying genetics of stroke and its comorbidities.

I have tried to address the concerns raised by the reviewers, which ideally doesn’t impact the original manuscript. Statistical limitation has been commented pertaining to P-values, which has been clarified here. However, certain minor concerns such as abbreviations have been resolved in the revised manuscript. My response to weakness and reviewer’s comments are mentioned below.

Reviewer #1 (Public Review):

Strengths:

The data provided here will provide a foundation for a lot of future research into the causes of the observed correlations as well as whether the observed differences in comorbidities across regions have clinically relevant effects on risk management.

Weaknesses:

As with any cross-national analysis of rates, the data is vulnerable to differences in classification and reporting across jurisdictions.

GBD data is the most robust and most comprehensive data resource which has been used and accepted globally in predicting the health metrics statistics.

GBD data indeed considers normalisations, regarding classification and reporting.

To the best of our knowledge this is the best available resource to consider all health metrics analysis.

Furthermore, given the increased death rate from COVID-19 associated with many of these comorbid conditions and the long-term effects of COVID-19 infection on vascular health, it is expected that many of the correlations observed in this dataset will shift along with the shifting health of the underlying populations.

I must clarify that we have used data prior to COVID-19.

But yes the patterns after COVID19 will shift due to the impact of covid. This makes the study even more relevant as the comorbid conditions of stroke are also the risk drivers for COVID19 and mortality. This shift has been reported by some authors, which has been discussed in the discussion.

Therefore, understanding the genetic factors underlying stroke and its comorbid conditions might help in resolving how COVID19 might differentially impact on health outcome.

We did highlight this aspect of COVID19 even in our original version.

Introduction 1st para:

“It is the accumulated risk of comorbid conditions that enhances the risk of stroke further. Are these comorbid conditions differentially impacted by socio-economic factors and ethnogeographic factors. This was clearly evident in COVID era, when COVID-19 differentially impacted the risk of stroke, possibly due to its differential influence on the comorbidities of stroke.”

Discussion 3rd para:

“Studies reported reduction in life expectancy in 31 of 37 high-income countries, deduced to be due to COVID-191 . However, it would be unfair to ignore the comorbid conditions which could also be the critical determinants for reduced life expectancy in these countries.”

Recommendations For The Authors:

On page 5, the authors make a note about Africa and the Middle East having the highest ASMR for high SBP and comment about the relative populations of these regions. The populations of the regions are irrelevant to the rate.

Since the study is on comorbid factors of stroke and its impact on mortality therefore, relative burden seems critical. This has been further elaborated here to justify the comment, which indeed is an integral part of the original manuscript.

Paragraph referred – Results section 2:

“Ethno-regional differences in mortality and prevalence of stroke and its major comorbid conditions

We observed interesting patterns of ASMRs of stroke, its subtypes and its major comorbidities across different regions over the years as shown in figure 1a, table 1 and supplementary files S2 & S3. When assessed in terms of ranks, high SBP is the most fatal condition followed by IHD in all regions, except Oceania (OCE) where IHD and high SBP swap ranks. Africa (AFR; 206.2/100000, 95%UI 177.4-234.2) and Middle East (MDE; 198.6/100000, 95%UI 162.8-234.4) have the highest ASMR for high SBP, even though they rank as only the third and sixth most populous continents (fig. S2), respectively.”

On page 17, the authors are alarmed by a large ratio between prevalence rates and mortality rates for certain conditions. This is confusing since this indicates that these conditions are not as dangerous as the other conditions.

This has been further elaborated here to justify the comment, which indeed is an integral part of the original manuscript.

Paragraph referred – Discussion para 1:

“While the global stroke prevalence is nearly 15 times its mortality rate, prevalence of comorbid conditions such as high SBP, high BMI, CKD, T2D are alarmingly 150- to 500-fold higher than their mortality rates. These comorbid conditions can drastically affect the outcome of stroke.”

In Figure 4, the colors are not defined.

In Structure plot colours are assigned as per each K, it doesn’t directly refer to any population. But the plot distinguishes the stratification of populations as per K. Ramasamy, R.K., Ramasamy, S., Bindroo, B.B. et al. STRUCTURE PLOT: a program for drawing elegant STRUCTURE bar plots in user friendly interface. SpringerPlus 3, 431 (2014). https://doi.org/10.1186/2193-1801-3-431

Reviewer #2 (Public Review):

Strengths:

The idea is interesting and the data are compelling. The results are technically solid.

The authors identify specific genetic loci that increase the risk of a stroke and how they differ by region.

Weaknesses:

The presentation is not focused. It would be better to include p-values and focus presentation on the main effects of the dataset analysis.

I presume the comment is made with reference to results with significant p-values.

P-values are mentioned in the main text when referring to significant decrease or increase with respect to global rates and time e.g. P-values for comparison of a year 2019, are based on regional rates to global rates of 2019. Supplementary table S2a (mortality) and S3a (prevalence) e.g. P-values for comparison between year is based on 2019 rates to 2009 rates in Supplementary table S2b (mortality) and S3b (prevalence) e.g. P-values for proportional mortality and proportional prevalence in Supplementary table S4 and S5 is also based on global rates.

Recommendations For The Authors:

It would be better to minimize the use of acronyms. Often one has to go back to decipher what the acronym stands for. It is fine to have acronyms in figure legends, if necessary. However, at least for regions, please do not use acronyms.

In the revised version we have tried to minimise the Acronyms.

Removed the acronyms for regions and other places wherever possible however, the diseases acronyms have been maintained as per the GBD terms.

Please focus the presentation on the main results. Currently, the presentation wanders and repeats itself a lot.

Since the manuscript tries to address the global and regional rates of prevalence, mortality and its relationship to genetic correlates, it is difficult not to repeat the same to stress the significant observations coming out of different analysis methods. This might reflect on some amount of repetitiveness but the intention was to stress the significant observations.

I would also recommend acknowledging and discussing socioeconomic factors earlier in the manuscript.

Current mention happens in 3rd para of Discussion

“The changing dynamics of stroke or its comorbid conditions can be attributed to multitude of factors. Often global burden of stroke has been discussed from the point of view of socio-economic parameters. Studies indicate that half of the stroke-related deaths are attributable to poor management of modifiable risk factors 8,9. However, we observe that different socio-economic regions are driven by different risk factors.”

Version 2

Abstract

Abstract

Burden of stroke differs by region, which could be attributed to differences in comorbid conditions and ethnicity. Genomewide variation acts as a proxy marker for ethnicity, and comorbid conditions. We present an integrated approach to understand this variation by considering prevalence and mortality rates of stroke and its comorbid risk for 204 countries from 2009 to 2019, and GWAS risk variant for all these conditions. Global and regional trend analysis of rates using linear regression, correlation and proportion analysis, signify ethnogeographic differences. Interestingly, the comorbid conditions that act as risk drivers for stroke differed by regions, with more of metabolic risk in America and Europe, in contrast to high SBP in Asian and African regions. GWAS risk loci of stroke and its comorbid conditions indicate distinct population stratification for each of these conditions, signifying for population specific risk. Unique and shared genetic risk variants for stroke, and its comorbid and followed up with ethnic specific variation can help in determining regional risk drivers for stroke. Unique ethnic specific risk variants and their distinct patterns of Linkage Disequilibrium further uncover the drivers for phenotypic variation. Therefore, identifying population and comorbidity specific risk variants might help in defining the threshold for risk, and aid in developing population specific prevention strategies for stroke.

Graphical Abstract

Introduction

Stroke affects over 101 million people worldwide and is ranked the second most fatal disease in the world, with 6.5 million deaths in 20191. Comorbid conditions of stroke are critical contributors to burden of stroke and the duration of the comorbid conditions can further determine the severity of stroke risk or mortality. Prevalence for comorbid conditions range from 43% to 94% and estimates can go as high as 99% above 66 years of age2. Prevalence and mortality risk in stroke has often been evaluated from socio-economic viewpoint, but it is also critical to understand the differences in drivers such as comorbid conditions. It is the accumulated risk of comorbid conditions that enhances the risk of stroke further. Are these comorbid conditions differentially impacted by socio-economic factors and ethnogeographic factors. This was clearly evident in COVID era, when COVID-19 differentially impacted the risk of stroke, possibly due to its differential influence on the comorbidities of stroke.

Mortality in Stroke, its subtypes and their comorbid conditions have a strong ethnic bias3,4,5. Genetics act as surrogate markers for ethnogeographic indices. It is important to understand which comorbid conditions are influenced by socio-economic indices, and how they impact the risk of stroke and their underlying genetic basis. A Danish study reported the effect sizes of association with comorbid conditions for stroke to have 15% higher mortality risk in presence of diabetes mellitus with end-organ damage, 20% for peripheral vascular disease, 25% for chronic pulmonary disease, 35% for congestive heart failure and atrial flutter, 45% for moderate to severe renal disease, and 1.8– to 2.4-fold for mild to severe liver disease6. A UK Biobank study on stroke multimorbidity reported 1.5x higher risk of mortality in those with 2 additional comorbidities and a ≈2.5× higher risk of mortality in those with ≥5 comorbidities over seven years7. Thus, the differential impact of comorbid or multimorbid conditions contributing to the additive effect of illness burden needs to be addressed from an ethnogenetic perspective. Devising an appropriate strategy for prevention of stroke burden, needs a careful evaluation of the underlying genetic signature for each of these comorbid conditions and distinguishing their ethnic bias.

The objective of the study was to understand what determines the differences in stroke burden around the globe. Variations in burden of stroke could be influenced by comorbid conditions, and incidentally both stroke and its comorbid conditions can be influenced by ethnogeographic factors and genetics can act as a stable proxy marker for all. To resolve this, we considered the prevalence and mortality of a total of eleven disease conditions, consisting of stroke and its comorbid conditions, across different continents and ethnicities from 2009 to 2019. The disease conditions were further stratified as per their ethnogeographic locations and their genetic risk variants extrapolated from GWAS data. This study would provide insights on the regional patterns of the burden of stroke and its comorbid conditions, and help in resolving it from an ethnic and genetic viewpoint. These insights would further aid in developing and strategizing regional and ethnic specific needs for prevention of the risk of comorbid conditions and stroke.

Results

Global mortality, incidence and prevalence rates of stroke and its comorbid conditions

Globally, stroke ranks as the second most fatal disease in 2019 (84.2/100000, 95%UI 76.8-90.2) among the eight diseases analyzed, preceded by ischemic heart disease (IHD; 117.9/100000, 95%UI 107.8-125.9) as shown in Figure 1-figure supplement 1 and Table S1. High systolic blood pressure (high SBP) ranks as the most fatal comorbid disease condition with an age-standardized mortality rate (ASMR) of 138.9/100000 [95%UI 121.3-155.7] among all conditions. Within stroke subtypes, ischemic stroke ranks highest globally with ASMR of 43.5/100000 [95%UI 39.08-46.8], followed by intracerebral hemorrhage (ICH) at 36.0/100000 [95%UI 32.9-38.7] and subarachnoid hemorrhage (SAH) at 4.7/100000 [95%UI 4.1-5.2]. The ranking of the ASMRs of the diseases follow the same trend throughout the last decade with minor exceptions, where high body-mass index (high BMI) improved its rank in 2014 by swapping with high low-density lipoprotein cholesterol (high LDL), and type 1 diabetes (T1D) dethroned chronic kidney disease (CKD) in 2019.

Regional (a) mortality rates and (b) prevalence rates for Stroke, its subtypes and comorbid factors.

The figure shows the age-standardized mortality and prevalence rates per 100,000 people for Stroke, its subtypes and its comorbid factors in 2009, 2014 and 2019 in the geographical regions. The size of the points indicates the rate and position indicates rank.

Global trends in crude and age-standardized incidence rates (ASIRs) show that stroke incidence ranks fourth, while IHD is on top followed by T2D and CKD (Figure 1-figure supplement 1, Table S1). Crude incidence rates of stroke and its subtypes increased in the last decade but ASIRs decreased, with exception of ischemic stroke, where an increase in ASIR was observed. Among other diseases, IHD, type 2 diabetes (T2D) and T1D show a continuous increase in the last decade, with T2D (280.1/100000, 95%UI 258.8-303.9) surpassing IHD (274.0/100000, 95%UI 242.9-306.4) in 2019 in crude rates. The global trend between crude and age-standardized prevalence rates (ASPRs) revealed that the ranking of stroke remains at sixth position throughout the years (Figure 1-figure supplement 1, Table S1). Among stroke subtypes, ischemic stroke ranks highest. The comorbid conditions, high SBP and high BMI, ranks first and second followed by CKD and T2D, and interestingly all rank above stroke. We also observe that though there is a continuous increase in prevalence rates in the last decade, the ranking of stroke or its comorbid conditions does not change over the years, with the exception of T1D ASPRs overtaking ICH. We were keen to understand if these global trends of ASMR and ASPR are influenced by region and ethnicity.

Ethno-regional differences in mortality and prevalence of stroke and its major comorbid conditions

We observed interesting patterns of ASMRs of stroke, its subtypes and its major comorbidities across different regions over the years as shown in Figure 1a and Figure 1-figure supplement 2, Table 1 and S2. When assessed in terms of ranks, high SBP is the most fatal condition followed by IHD in all regions, except Oceania where IHD and high SBP swap ranks. Africa (206.2/100000, 95%UI 177.4-234.2) and Middle East (198.6/100000, 95%UI 162.8-234.4) have the highest ASMR for high SBP, even though they rank as only the third and sixth most populous continents (Appendix Figure 1), respectively. Both high SBP (−0.64% to –2.25% EAPC) and IHD (−0.45% to –1.17% EAPC) show a decreasing trend for ASMRs in all regions. However, only Europe shows a significant decrease for high SBP (−2.26%; P=0.009) and IHD (−2.37%; P=0.006) in the decade. Stroke has a decreasing trend for ASMRs with East Asia and Europe showing a significant decrease of –2.2% (P=0.021) and –2.6% (P=0.03), respectively. Stroke has the highest mortality in East Asia (127.1/100000, 95%UI 104.9-150.5 in 2019), and is the only region that ranks stroke higher than IHD. Though Europe, Middle East and Central & South Asia have ASMRs similar to global rates for stroke, Central & South Asia ranks stroke as the third most fatal factor, while America, Europe and Middle East ranks it fifth. Oceania (62.1/100000, 95%UI 34.1-90.2) and America (40.3/100000, 95%UI 36.2-43.1) have lowest rates for stroke in 2019.

Among the stroke subtypes, ICH and ischemic stroke show maximum ethnogenetic differences in mortality rates (14.1/100000 to 61.4/100000) and ranking (4th to 9th) in 2019. While ICH shows a significant decrease in ASMR in East Asia (−3.53%, P=0.009), ischemic stroke shows a significant decrease in Europe (−2.37%, P=0.06). High BMI and high LDL rank in the top five but their mortality rates differed across all regions, with the highest rates for both in the Middle East. Only Europe shows a significant decrease in high LDL (−2.53%, P=0.03) over the decade. The Middle East has the highest ASMRs due to IHD and high SBP, followed by Africa, Central & South Asia, East Asia and Europe, all having rates higher than global. All continents have similar mortality rates for T2D and CKD across the years, except Oceania, where the T2D rate is nearly three times CKD rate. Africa has the highest mortality rate for T1D (1.59/100000, 95%UI 1.2-1.9).

ASPRs also showed an interesting pattern of distribution, and, in contrast to mortality, showed an increase over the decade (Figure 1b and Figure 1-figure supplement 3, Table 2 and S3). Highest ASPRs were observed for high SBP across all regions, except America, Middle East and Oceania, where high BMI has most prevalence. While EAPC of prevalence of high SBP showed significant decrease in all regions (−0.38% to –1.77%), except Central & South Asia, high BMI showed a significant increase (1.77% to 5.6%) in all (Table 2). The ASPR ranking of CKD and T2D rose to top five, in sharp contrast to their ASMR rankings. Prevalence of CKD (EAPC 0.24% to 0.7%) and T2D (EAPC 0.6% to 2.18%) is significantly increasing in all regions. For all other diseases, the pattern of ranking and rates across regions were stable with minor exceptions. Stroke ranks sixth for ASPRs in all regions, and it is interesting to note that ASPRs of all the comorbid conditions, except T1D, rank above stroke. While Europe shows a significant decline in ASPRs of stroke (−0.9%, P=0.008) and ischemic stroke (−0.85%, P=0.03) across the years, East Asia shows a significant increase for stroke (0.7%, P=0.02) and ischemic stroke (1.09%, P=<0.001). Globally, T1D swapped its prevalence ranking with ICH in 2014, largely influenced by the significant increase in prevalence in Middle East (2.83%, P=<0.001). However, the highest prevalence of T1D is in Europe and Oceania. The prevalence of IHD has remained nearly constant in all continents in the last decade, except Oceania (−0.43%), America (−0.49%) and Middle East (−0.27%) which shows a significant decrease. In 2019, Middle East has the highest prevalence for IHD (4843.02/100000, 95%UI 4243.02-5442.58), while America (1695.6/100000, 95%UI 1530.8-1871.9) has the lowest ASPR, less than half the rate of the Middle East. We were further keen to understand if these regional differences in mortality and prevalence rates also reflect a socio-economic bias and if so, does it reflect in a category of comorbid conditions.

Contribution of metabolic risk and hypertension in stroke based on ethnogenetic locations

When the prevalence and mortality rates of stroke and its comorbidities were grouped into three groups, namely strokes, metabolic disorders and high SBP, we find that out of the three proportional mortalities shown in figure 2, strokes group has the highest proportion (37.1%-47.2%) across all years and regions, except Oceania and America, where instead, metabolic disorders have the highest proportion (39.1%-42.2%) that is significantly higher compared to global proportion (Table S4). East Asia has the highest proportional mortality for strokes among all regions in all three years (44.7%-47.2%). This was in sharp contrast to the prevalence proportion of strokes (4.6%-9.1%), which was the least among the three groups, with the highest proportional prevalence for strokes being 8.9%-9.1% in Central & South Asia. The proportional mortality for high SBP (22.0%-30.5%) is very similar across the regions. Metabolic disorders have significantly higher proportional prevalence compared to global in Middle East, Oceania and America, the highest being in America (63.9%, Table S5). Asian and African regions have lowest proportional prevalence for metabolic disorders, with Central & South Asia having significantly lower proportions. However, these regions have the highest prevalence proportion for high SBP (46.6% – 54.3%), with Central & South Asia having a significantly higher proportion compared to global. On the other hand, the Middle East, Oceania and America have significantly lower proportional prevalence compared to global. We were further keen to understand the correlation among ASMR and ASPR of comorbid conditions among ethnogeographic regions.

Proportional (a) mortality rates and (b) prevalence rates for Strokes, High SBP and Metabolic disorders.

The figure shows the proportional mortality and prevalence for all strokes in yellow (ischemic stroke, intracerebral hemorrhage, subarachnoid hemorrhage and ischemic heart disease), high systolic blood pressure in red, and metabolic disorders in green (high BMI, high LDL, diabetes mellitus type 1 and 2, chronic kidney disorder). CSA – Central and South Asia, AFR – Africa, EAS – East Asia, EUR – Europe, MDE – Middle East, OCE – Oceania, AMR – America.

Correlation among prevalence and mortality rates based on ethnogeographic region

Correlation between ASMRs and ASPRs for stroke and all comorbid conditions across each ethnogeographic location is shown in figure 3. High SBP prevalence and mortality show a strong positive correlation in Central & South Asia but a strong negative correlation in the Middle East. The prevalence and mortality of high BMI has a strong negative correlation in Central & South Asia and Middle East populations, but a strong positive correlation in East Asia. Prevalence of CKD has negative correlation with mortality rates in East Asia. Prevalence of T2D has negative correlation with mortality rates in Oceania and positive correlation in America. It is interesting to note that there was not much of a correlation in mortality and prevalence rates for most of the conditions. For overall stroke, though minor correlations between various ethnicities are seen, this becomes alarmingly clear in the stroke subtypes. The correlation matrix of mortality and prevalence rates of stroke and its comorbidities does reflect strong ethnogeographic distinctions, which formed the basis of further investigation on the genetic basis of stroke and its comorbidities.

Correlation of mortality versus prevalence rates of Stroke, its subtypes and comorbid factors.

The figure shows the Pearson correlation coefficients of age-standardized mortality rates versus prevalence rates across different continents for 2014. CSA – Central and South Asia, AFR – Africa, EAS – East Asia, EUR – Europe, MDE – Middle East, OCE – Oceania, AMR – America.

Ethnogeographic stratification of stroke and its comorbidities based on GWAS data

To resolve the ethnogeographic distinctions for stroke and its major comorbid conditions based on their genetic risk, we considered all GWAS loci for stroke and its major comorbid conditions, and subjected it to stratification analysis. From the GWAS loci, we observed a distinct population structure that distinguished ethnogeographic populations based on their genetic signatures (Figure 4). For all diseases, except high BMI, the individuals clustered into five groups, each corresponding with the five super-populations from 1000 Genome project namely, African, East Asian, South Asian, European and American. For high BMI, the individuals clustered into three groups corresponding to African, East Asian, European. Though broad clustering of ancestral populations among the diseases looks similar, the proportions of ancestral populations in certain diseases vary greatly. Among Stroke, IHD and T2D risk variants, the populations structured in a similar way, while for T1D, CKD and LDL the patterns were slightly different in European and South Asian ancestry. Whereas for high SBP the major fluctuations were observed in South Asian and East Asian populations. This stratification was further explored using population-based clustering, where a similar pattern was observed in PCA plots for stroke and its comorbid conditions (Figure 5). African population seems to be a distinct outlier for most diseases, and the East Asian comes a distant second in the cluster pattern.

Population structure of risk variants for Stroke and its comorbid factors using a model-based clustering.

The proportion of ancestral populations was estimated from the genotype of risk variants of each disease in unrelated individuals from the 1000 Genomes project using a model-based clustering approach. The individuals were represented by their super-population in 1000 Genomes (African, East Asian, South Asian, European and American). The estimated ancestral populations clustered into either 5 clusters or 3 clusters (in case of BMI) represented by the different colors.

Clustering of risk variants for Stroke and its comorbid factors using a PCA.

The eigenvectors were estimated from the genotype of risk variants of each disease in unrelated individuals from the 1000 Genomes project using PCA clustering. The individuals were represented by their super-population in 1000 Genomes (African, East Asian, South Asian, European and American) shown in five different colors.

Similar patterns were observed in stratification analysis after grouping the GWAS loci of stroke and its comorbidities into three groups as before, namely strokes, metabolic disorders and high SBP. The African, East Asian and South Asian populations had distinct structure in all three groups (Figure 4-figure supplement 1). These observations do indicate that the underlying genetic factors of stroke and its comorbid factors can be the real indicators of ethnogeographic patterns of risk for stroke and its comorbid conditions. However, we were further keen to understand the extent of shared and unique individual risk variants across stroke and its comorbid condition and how these unique or shared variants can help in distinguishing their relevance across ethnicities.

Shared and unique risk variants of stroke among the different ethnogenetic regions

The unique and shared individual variants across stroke and its comorbid conditions were identified from the GWAS data irrespective of ethnicity. We find the majority of the risk variants were unique to a disease condition, however, several risk variants were also seen to be shared with stroke and other comorbid conditions as seen in Figure 6. We were further keen to have a deeper insight into the distribution of risk variants in stroke across ethnicities. Stroke has only 55% of the risk variants common to all the five populations as seen in Figure 7 and Table S6. Two groups of populations share the most number of variants, namely, the Africa-America-Europe-South Asia (6% of variants shared) group and the East Asia-America-Europe-South Asia (4%) group. Africa has the highest number of unique variants for stroke (6%), followed by Europe (3%).

Distribution of unique and shared risk variants in stroke and its comorbid conditions.

The barplot shows the number of GWAS risk variants unique for stroke and its comorbidities, as well as number of GWAS risk variants shared among the different diseases. The intersection of diseases is indicated below. Intersections containing stroke are highlighted in red. The total number of GWAS risk variants considered are stroke (366), IHD (1137), High SBP (419), T2D (2227), T1D (269), CKD (125), High BMI (27) and High LDL (1734).

Risk variants of Stroke shared among super-populations.

The GWAS risk variants for stroke (total 291) shared among the five super-populations in 1000 Genomes (African, East Asian, South Asian, European and American). A risk variant was considered to be present in a population if the alternate allele frequency in 1000 Genomes was greater than or equal to 0.05.

South Asia has two unique variants for stroke, rs528002287 and rs148010464, which maps to genes PCSK6 and the intergenic region of PLA2G4A/LINC01036, respectively. The variants rs528002287 and rs148010464 are low frequency variants in South Asia with a MAF of 0.053 and 0.054, respectively. We were further keen to understand how these unique variants in South Asia are tagged to the nearby variants of different frequencies in the different ethnicities. Can the LD patterns of rare allele and common allele help in distinguishing the ethnogeographic distinction in phenotype variation. In comparison, the linkage disequilibrium (LD) pattern of low frequency variants and common variants across ethnicities showed contrasting patterns. The LD plots between the unique variants and low frequency variants tagged to it clearly demonstrated a unique LD pattern in South Asia, compared to other populations (Figure 8). Contrastingly, the LD plots of common variants tagged to the risk variant of stroke unique to South Asia showed similar LD patterns among all populations (Figure 8-figure supplement 1). These differences might also reflect unique or distinct phenotypic differences among ethnicities for risk in stroke.

Linkage disequilibrium of low frequency variants proxy to risk variants of stroke unique in South Asian population.

Proxy variants with frequency less than 0.1 in 1MB region flanking the stroke risk variants unique in South Asia (a) rs528002287 and (b) rs148010464. The risk variants are marked in blue box. LD plots of variants in different 1000 Genome super-populations (AFR-Africa, EAS-East Asia, SAS – South Asia, AMR – America, EUR – Europe) are shown.

Discussion

To the best of our knowledge, this is the first study that explored the burden of stroke and its comorbid conditions across regions, stratifying and distinguishing their unique features based on their genetic background extrapolated from the GWAS risk loci. The dynamics of different rates of stroke, its subtypes and comorbid factors do reflect ethnogeographic differences. Globally, the prevalence and incidence rates of stroke have increased, while mortality rates decreased with minor shifts in ranking in the last decade. Interestingly, the incidence and prevalence rank of stroke rates remain the same globally, but for mortality it ranks third, preceded only by IHD and high SBP. While the global stroke prevalence is nearly 15 times its mortality rate, prevalence of comorbid conditions such as high SBP, high BMI, CKD, T2D are alarmingly 150– to 500-fold higher than their mortality rates. These comorbid conditions can drastically affect the outcome of stroke. Interestingly, these disparities in rates get further widened when evaluated from an ethnogeographic perspective. The age-standardized prevalence of stroke in 2019 ranges from lowest in Central & South Asia (858.5/100000, 95%UI 737.6-979.4) to highest in East Asia (1513.1/100000, 95%UI 1390.7-1635.5), in contrast to mortality rates, lowest in America (40.3/100000, 95%UI 36.2-43.1) and highest in East Asia (127.7/100000, 95%UI 104.9-150.5).

The rates of stroke, its subtypes and comorbid conditions do correlate to some extent but their ranking varies significantly. In terms of ranking, stroke ranks sixth in prevalence across all ethnogeographic locations, but ranks second in mortality in East Asia and third in Central & South Asia and Africa. We find that among the considered comorbid conditions, some rank above stroke in both incidence and prevalence, however in terms of mortality, stroke ranks highest with exception to high SBP. Similarly, the ranking of the comorbid conditions also varies when the global population is stratified based on ethnogeographic locations. In the last decade, there has been tremendous development in the healthcare industry globally, but this is not reflected in the mortality or prevalence data from 2009-2019. Ranking of comorbid conditions by the rates is very crucial to identify ethnic-specific comorbid risk that can be helpful in guiding and managing stroke risk.

The changing dynamics of stroke or its comorbid conditions can be attributed to a multitude of factors. Often the global burden of stroke has been discussed from the point of view of socio-economic parameters. Studies indicate that half of the stroke-related deaths are attributable to poor management of modifiable risk factors8,9. However, we observe that different socio-economic regions are driven by different risk factors. Considering that Europe, America, Oceania and the Middle East represent high socio-economic regions, the comorbid conditions that drive prevalence and mortality rates here seem to be more of metabolic in nature, while for South Asians, high SBP is the prominent factor. It is evident from the correlation between prevalence and mortality for stroke, its subtypes and comorbid conditions that there is an ethnic and comorbid specific correlation, which possibly does not reflect a clear socio-economic distinction. The comorbid conditions for stroke subtypes also differ to a large extent. Diabetes is a comorbid condition for stroke but not for SAH, and this risk is ethnic-specific10,11. Therefore, it is very pertinent to understand the stroke risk from an ethnic view point, beyond the boundaries of socio-economic criteria, as the drivers of comorbid risk and ethnicity rely on genetic and epigenetic components. Studies reported reduction in life expectancy in 31 of 37 high-income countries, deduced to be due to COVID-1912. However, it would be unfair to ignore the comorbid conditions which could also be the critical determinants for reduced life expectancy in these countries.

Stroke has a complex etiology, which is further influenced by its comorbid conditions and this impacts its phenotypic variability. A strong genetic risk drives both stroke and its comorbid conditions. Genetic risk variants for diabetes, cardiovascular disease, diabetic retinopathies and nephropathies, hypertension, inflammation and kidney diseases have been reportedly shown to have strong ethnogenetic variation13. Implications of ethnogenetic differences were evident when GWAS genes for stroke and all studies of comorbid conditions were used to stratify the 1000 genome super-populations. We observed that the GWAS risk loci for stroke and its comorbid conditions like high BMI, high LDL, high systolic BP, type 2 diabetes, type 1 diabetes, and chronic kidney disease could stratify the super-populations based on its ethnogenetic considerations. Fluctuations in genetic structure of stroke and its comorbid conditions signify the impact of ethnic variations on mortality and prevalence rates. Stroke accounts for approximately 20% of deaths in diabetics14,15. Diabetics and pre-diabetics, and the duration of diabetes have been reported to have increased risk of stroke, which gets aggravated in African-Americans14,15. As stroke and its comorbid conditions are heavily influenced by lifestyle, high-income countries showed evidence of metabolic disorders being the major cause of concern for both prevalence and mortality. Similar observation in a UK biobank cohort study on stroke suggests that genetic and lifestyle factors were independently associated with incident stroke, which emphasizes the benefit of entire populations adhering to a healthy lifestyle, independent of genetic risk16.

Metabolic risk variations could also be a reflection of their underlying genetic differences. Significant differences among ethnicities in metabolism of various macromolecules have been reported17. Genetic variations demonstrate inter-ethnic differences in LDL levels resulting in differential impact on dyslipidemia18. A meta-analysis on European, East Asian and African-American ethnicities revealed that common variants of CDH13 and ADIPOQ regulate adiponectin levels, an important component of BMI indicator19. ALDH2*504Lys allele has been reported to be associated with high BMI, increased tolerance of alcohol, high SBP, and decreased high-density lipoprotein in East Asians20,21. MEGASTROKE consortium on stroke and its comorbid factors reported five variants associated with blood pressure, two with LDL cholesterol and reported that all stroke subtypes were associated with a Genetic Risk Score for high SBP22. A study on IHD, T2D in European populations with different health-care systems, and local population substructures reported Polygenic risk score with similar accuracy across Europeans, to a lesser extent to South and East Asian populations, and very poor transferability for Africans23.

Even stroke subtypes show a strong ethnic variation22,24. While COL4A1 and COL4A2 were common denominators for most of the stroke subtypes, high-density lipoprotein was inversely associated with small vessel stroke25. The INTERSTROKE study on stroke subtypes reported that among comorbid factors, high SBP was significantly associated with ICH and diabetes, cardiac issues, and apolipoproteins with ischemic stroke26. Ischemic stroke associated with ApoB/ApoA1 ratio had higher (67.6%) population attributable fraction (PAF) in Southeast Asia compared to Western Europe, North America, and Australia (24.8%), while ischemic stroke associated with atrial fibrillation had lower PAF in South Asia (3.1%) compared to the rest (17.1%)26. Emerging studies like novel ethnic specific genetic variants, such as SUMOylation pathway in Indians27, SFXN4 and TMEM108 in Africans,28 indicate the involvement of different pathways among different ethnicities in stroke. The A allele of c.*84G>A loci in CETP gene was found to be a risk factor for IHD in South Asians29. High SBP was found to be a risk factor in all major stroke subtypes except lobar ICH30. Therefore, identifying differential risk in different ethnicities for stroke and its subtypes, and its impact on comorbid conditions might also indicate different treatment modalities which can minimize adverse metabolic side effects.

Identifying the pattern of genetic variation is critical in distinguishing stroke and its endophenotypic variations. The risk variants rs528002287 (locus 15q26.3) in PCSK6 and rs148010464 (locus 1q31.1) an intergenic variant in PLA2G4A/LINC01036 for stroke were unique to South Asia, and were found to be associated with cardioembolic stroke and small vessel stroke in South Asians27. A recent INTERSTROKE study reported the association of short sleep duration with increased risk for stroke to be highest in the South Asian ethnicity (OR 9.13, 95% CI 5.86-14.66)31. Decreased sleep quantity and quality has been reported to increase blood pressure32, prevalence for which was found to be highest for South Asians in our study. These observations are interesting as PCSK6 is known to regulate sodium homeostasis and thereby maintain diastolic blood pressure.33,34 Reports also indicate 2.34 fold difference in the expression of PCSK6 during maintenance phase of hypertension35. PCSK6 has also been reported to be involved in processing of precursors of Melanin Concentrating hormone (MCH) under certain conditions36. MCH is known to play a central role in promoting and stabilizing sleep37,38. In insomnia patients, PLA2G4A was reported to be upregulated by 1.88 fold after improvement in sleep39. In sleep deprived mice, glycolytic pathway and lipid metabolism was upregulated and expression of PLA2G4A was downregulated40. These observations are interesting as PLA2G4A is known to play a role in the metabolism of phospholipids, production of lipid mediators and the release of arachidonic acid41,42. Arachidonic acid is involved in signaling pathways of metabolic processes like release of insulin and glucose disposal43–45. PLA2G4A also plays a role in the production of pro-thrombotic TXA2 and thus, inhibition of PLA2G4A can reduce platelet aggregation and thromboembolism46. Thus, the risk of unique genetic variants in PCSK6 and PLA2G4A in South Asian ancestry may indicate a unique endophenotype for stroke, which might also indicate the influence of underlying risk variants for comorbid conditions, e.g. PLA2G4A in metabolic processes. GWAS has yielded numerous common risk alleles that are associated with various human phenotypes47,48. Since the rationale for GWAS is the ‘common disease, common variant’ hypothesis, it has been able to identify only those common variants with a moderate effect on the associated trait and thus, these identified variants only explain a small proportion of the heritability of the trait. One of the major conclusions from the 1000 Genomes Project was that most variations in the human genome are rare and unique to specific sub-populations49,50. From an evolutionary point of view, alleles with strong effects that are detrimental will be controlled by purifying selection keeping its frequency low. Hence, rare and low frequency variants might be variants with large effects on traits51. Examples of genes like ABCA1, PCSK9 and LDLR52,53, which carries both common variants with moderate effects as well as rare variants with large effect for lipid levels indicate that genes can contain both types of variants associated with a complex trait. Using this logic, we were keen to identify the influence of common and rare variants in stroke and its comorbid conditions and their pattern of LD in distinguishing ethnic specific risk.

While we observe the majority of the GWAS variants associated with stroke are common variants, a minority of these are low frequency variants with below 10% frequency. The alternate alleles of these variants were found to be present only in specific super-populations of 1000 Genomes, while the variants are monoallelic in the other super-populations. This difference among populations gets further enhanced when we look at the LD patterns of these low frequency variants with other low frequency variants which are in proxy. Distinct LD blocks with these unique rare variants present in one super-population were seen to be absent in other populations. On the other hand, LD patterns of common variants (frequency greater than 10%) in the same regions show similar LD patterns for all populations. Thus, the rare variant hypothesis could explain a significant proportion of the differences in burden of stroke seen across populations. Such genes identified could be possible candidate genes for identifying rare and low frequency variants that could play a role in the heritability of stroke and its comorbid factors.

Conclusions

The dynamics of incidence, prevalence and mortality rates of stroke and its subtypes along with its comorbid risk factors reflect strong ethnogeographic differences. Our work highlights that these ethnogeographic differences for stroke and its comorbid conditions need to be evaluated and stratified based on their ethnogenetic background. Genetic variables should be considered as primary evidence as they define the threshold for biochemical, metabolomic or epigenetic variables. The different socio-economic regions are driven by different risk factors of stroke and low frequency variants could be playing a role in the differences in burden of stroke seen in the different regions. Identifying population specific unique variants for stroke and its comorbid conditions might refine the drivers for endophenotypic variations for stroke risk. We would like to suggest that integrating public health genomics and articulating it with comorbid conditions for stroke should be considered crucial irrespective of the economic status, as both lower and higher socio-economic regions have different drivers of stroke risk.

Methods

Data sources

We obtained age-standardized incidence rates (ASIRs), age-standardized prevalence rates (ASPRs) and age-standardized mortality rates (ASMRs) for a total of eight diseases and three disease conditions in 204 countries, for the years 2009 to 2019 using the GBD Results Tool (vizhub.healthdata.org/gbd-results/)54 and from NCD Risk Factor Collaboration (NCD-RisC, ncdrisc.org/) study55,56 in November 2021. GBD codes for the selected diseases were B.2.2 (Ischemic heart disease, IHD), B.2.3 (Stroke), B.2.3.1 (Ischemic stroke), B.2.3.2 (Intracerebral hemorrhage, ICH), B.2.3.3 (Subarachnoid hemorrhage, SAH), B.8.1.1 (Diabetes mellitus type 1, T1D), B.8.1.2 (Diabetes mellitus type 2, T2D) and B.8.2 (Chronic kidney disease, CKD). Three disease conditions, which are also comorbid factors of stroke, were also selected, high systolic blood pressure (SBP) (>110-115 mmHg), high body mass index (BMI > 23.0 kg/m2), and high LDL cholesterol. For these, ASMRs in 204 countries, as well as global rates for 2009 to 2019 were obtained using the GBD Results Tool, and age-standardized prevalence percentages in 204 countries for the years 2009 to 2016 (data for 2017 to 2019 was not available) were obtained using NCD-Risc website (ncdrisc.org/)55,56. Global crude rates and age-standardized incidence, prevalence and mortality rates were obtained from GBD and NCD for all the diseases and disease conditions (Figure 1-figure supplement 1, Table S1). GBD 2019 and NCD-RisC study compiled with the GATHER Guidelines.

Spatio-temporal trend analysis and estimated annual percent change

The 204 countries were grouped into eight geographic regions namely Global, America, Europe, Middle East, Africa, Central Asia & South Asia, East Asia, and Oceania, based on ethnicity (Table S7, Appendix Figure 1). The prevalence percentages were converted into ASPRs for each disease condition. ASMRs and ASPRs for each region were obtained using Bayesian model averaging of linear regression models with Markov chain Monte Carlo sampling. BIC was the model selection criteria. Poisson distribution with global rates as lambda was the prior distribution on the models. The population of each country was used as weights. 10000 draws from the posterior distribution of model parameters were used to obtain the point estimates (mean of the draws) and 95% uncertainty intervals (2.5th to 97.5th percentiles of the posterior distributions) of mortality and prevalence rates for each region for years 2009 to 2019. Rate estimates presented in this paper are age-standardized rates per 100,000 population.

The ASMRs and ASPRs thus obtained for the different diseases in each region were subjected to a temporal rank analysis for 2009, 2014, and 2019 using custom perl scripts and change in the trend was plotted as a bump plot. The size and position of the points indicates the rate and rank respectively. To quantify the temporal trends, estimated annual percentage change (EAPC) was modeled using Poisson regression using a generalized linear model for the log-transformed rates: log(y)= β0+ β1×1 + β2×2 + ….+βpxp, where y is the age-standardized rate, xi are the calendar years, βi are the rate trends. Under the assumption of linearity of log of age-standardized rate with time, EAPC =100* exp(β^) – 1. The 95% uncertainty interval is calculated as CI(EAPC) = β + (Z(1-α)/2) x SE, where α is the confidence level, SE is the standard error of β^.

EAPC for ASMRs and ASPRs were calculated for the time period 2009 to 2019 (Table 1 and 2). EAPC for high SBP prevalence was calculated from 2009 to 2015 and for high BMI prevalence from 2009 to 2016. EAPC was considered statistically significant if the uncertainty interval of EAPC did not cross zero. Statistical significance of spatio-temporal difference in rates was calculated using chi-square test by assuming the rate to be under Poisson distribution. 2019 ASMRs (or ASPRs) in each region were compared with global ASMR (or ASPR) to compare change over locations and with 2009 ASMRs (ASPRs) to compare change over time. 2014 ASPRs were used when 2019 data was not available. P-values (two-sided) for ASMRs and ASPRs are shown in Table S2 and S3, respectively. Chi-square tests were done using Open Source Epidemiologic Statistics for Public Health (www.openepi.com). The relation between ASMRs and ASPRs of each region was measured using Pearson correlation. For correlation, we considered the data of 2014 due to the complete spectrum of data availability. All analysis, unless specified, was done using R Statistical Software (version 4.1.2)57 with packages BAS58, Rcan59, corrplot60, dplyr and ggplot2.

Proportional mortality and prevalence

The diseases were classified into three categories as stroke (ischemic stroke, ICH, SAH and IHD), metabolic disorders (High BMI, High LDL, T2D, CKD, and T1D) and high SBP. To determine the proportion of each category in a region, total ASMRs were scaled to 100 for all three years. The same was done for ASPRs (high LDL cholesterol data was not available). Statistical significance of difference in proportions was calculated using a one-sample test for binomial proportion using normal-theory method. The proportion mortality in each category was compared in a pairwise manner with global as well other regional proportions to calculate the two-sided p-value. The same was done for proportional prevalence. P-values for proportional mortality and prevalence are shown in Tables S4 and S5, respectively. Proportion comparisons were done using Open Source Epidemiologic Statistics for Public Health (www.openepi.com/Menu/OE_Menu.htm).

Population Structure Analysis

For evaluating the ethnogenetic perspective of stroke and its comorbid conditions we considered the risk variants associated with each disease. The risk variants were obtained from GWAS Catalog (www.ebi.ac.uk/gwas/home), during the period November 2021 to August 2022. The trait IDs used to retrieve data from GWAS Catalog for the different diseases were EFO_0000712 (Stroke), EFO_0001645 (IHD), EFO_0000537 (High SBP), MONDO_0005148 (T2D), MONDO_0005147 (T1D), EFO_0003884 (CKD), EFO_0007041 (High BMI) and EFO_0004611 (High LDL). The position of the risk variants in GRCh37 assembly was obtained, and variants less than 10000 bp apart were excluded using custom perl scripts. The total number of risk variants for each disease thus obtained are shown in Table S8. The genotype of identified biallelic autosomal SNPs in unrelated individuals was extracted from the 1000 Genome VCF files (www.internationalgenome.org)53. The proportion of ancestral populations in each individual was estimated from their genotype using a model-based clustering approach61. The admixture model with correlated allele frequencies was specified to cluster the individuals into either five clusters or three clusters (in case of BMI). The genotype was converted to eigenvectors using principal component analysis62,63.

Shared and Unique risk variants among ethnicities

The number of GWAS risk variants for stroke shared among, as well as, unique to the five super-populations in 1000 Genomes (African, East Asian, South Asian, European and American) was obtained (Figure 6 and Table S6). A risk variant was considered to be present in a population if the alternate allele frequency in 1000 Genomes was greater than or equal to 0.05. For each GWAS risk loci, the gene(s) the variant maps to was obtained and shared genes among the populations was also estimated (Table S6).

Calculation of Linkage Disequilibrium (LD) of risk variants of stroke in South Asia

Proxy variants (R2 > 0.01) in the region –/+ 500 Kb of the risk variant of stroke present in the South Asian population was obtained from LDlink64. Among the proxy variants, variants with minor allele frequency (MAF) less than 0.1 were selected along with the risk variant as low frequency variants, and variants with MAF greater than 0.1 were termed as common variants. LD between two alleles A and B is quantified using the coefficient of linkage disequilibrium DAB calculated using the equation DAB = pAB-pApB, where pi represents the frequency of the allele i or haplotype i. To be able to compare the level of LD between different pairs of alleles, D is normalized as follows: D’ = D/ Dmax, where Dmax = max{-pApB, –(1-pA)(1-pB)} when D < 0 and Dmax = min{pB(1-pA), pA(1-pB)} when D > 065. Estimates of D’ were calculated separately for low frequency variants and common variants, and plotted using the R library gaston66.

Supporting information